Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas

### 2021-04-27 analysis/paper/paper.Rmd
VARIOUS, MINOR EDITS

```{r setup, include = FALSE}
library(conflicted)
library(tidyverse)
conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")
library(lubridate)
conflict_prefer("month","lubridate")
library(glue)
library(DBI)
library(dbplyr)
library(sf)
library(flextable)
library(patchwork)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggspatial)
library(ggsci)
library(ggrepel)
library(gghighlight)
library(ggtext)
library(glmmLDTS)
library(mgcv)
library(ragg)

@@

The global climate crisis is causing considerable reduction in Arctic sea ice extent,
volume, and seasonal presence [@kwok2018erl; @meier2014rg; @wang2017dsrpitsio]. These
changes have ripple effects on Arctic ecosystems, taxa, and the human communities who live
in the region [@huntington2020ncc]. Such disruptions are a particular cause of concern for
ice-associated seals (ringed, bearded, spotted, and ribbon seals) that depend on spring
and early summer sea ice (March-June) in the Bering and Chukchi seas as a platform for
important functions, such as pupping, nursing, breeding, and molting [@boveng2009a;
@boveng2013; @cameron2010a; @kelly2010]. Limited data and large knowledge gaps prevent our
ability to make definitive predictions about the ultimate effects of changes in sea ice on
the behavior, abundance, and distribution of these seals. Knowledge about biological
constraints on the phenology of reproductive and molting behavior is generally lacking, so
it is difficult to predict the readiness with which ice-associated seal species will be
able to adapt to future changes (e.g., by adjusting pupping or molting schedules to
earlier dates or different locales). Additionally, trends in abundance of these species
are largely unknown, so it is difficult to say what effect, if any, declines in sea ice
habitat have had, or will have, on seal densities.

For ice-associated seals, haul out is tightly linked with sea ice. Ribbon seals rest out
of the water almost exclusively on sea ice and are mostly pelagic outside the spring
pupping, breeding, and molting season [@burns1981]. While spotted and bearded seals rest
on beaches during ice-free months, they strongly prefer sea ice as a resting platform
during the spring and early summer [@burns2009]. The remote nature of the Bering and
Chukchi seas mean direct observation of seal behavior is limited. Bio-logging devices are,
thus, especially useful tools for collecting key information on movement and haul-out
behavior for these species. Records of time spent out of the water provide valuable data
for identifying covariates that explain variation in haul-out behavior. For instance, in
the Antarctic, Bengtson and Cameron [-@bengtson2004pb] relied on bio-logging data to
demonstrate greater haul-out propensity in juvenile crabeater seals (*Lobodon
carcinophaga*) than adults, with highest probabilities in February and at times close to
solar noon. In the Arctic, Von Duyke et al. [-@vonduyke2020ee] demonstrated seasonal
changes between diurnal and nocturnal haul-out behavior of ringed seals. Bengtson et al.
[-@bengtson2005pb] documented a higher propensity for ringed seal basking near solar noon,
as did Ver Hoef et al. [-@verhoef2014sm] in an analysis of bearded (*Erignathus
barbatus*), ribbon (*Histriophoca fasciata*), and spotted (*Phoca largha*) seals using
much larger sample sizes. Olnes et al. [-@olnes2020pb] showed that the proportion of time
bearded seals spent hauled out progressively increased through spring and summer. And, Ver
Hoef et al. [-@verhoef2014sm] found haul-out probabilities increased gradually starting in
March and peaking in May and June for bearded, ribbon, and spotted seals. However, their
primary focus was to develop a correction factor for estimating seal abundance, and like a
subsequent analysis by Conn et al. [-@conn2014mee], they did not attempt to separately
analyze haul-out schedules for different age and sex classes (although initial models with
sex and age as fixed effects were insignificant at p = 0.05; [@verhoef2014sm]).

Several authors have used logistic regression-style analyses to estimate the probability
of sea ice use from data garnered from bio-loggers. In these models, haul-out
probabilities are expressed as a function of predictive covariates, such as time of day,
day-of-year, sex, age class, and environmental conditions (e.g., [@reder2003pb],
[@bengtson2004pb], [@bengtson2005pb], [@udevitz2009cjz], and [@verhoef2014sm]). However,
sample sizes have often been low and insufficient to permit strong inference about life
history and/or seasonal variation in haul-out probabilities. For instance, Bengtson and
Cameron's [-@bengtson2004pb] study included 5 adult and 2 juvenile crabeater seals, while
Bengtson et al.'s [-@bengtson2005pb] study was based on 6 telemetered ringed seals (*Phoca
hispida*) in the Chukchi Sea. These studies are often further limited by logistical
constraints on tagging fieldwork and the attachment duration or operational life of
satellite tags.

Ultimately, knowledge of trends in phenology and abundance (or life history surrogates
such as survival and recruitment) will be necessary to make credible quantitative
predictions about the effects of the climate crisis on the abundance and distribution of
Arctic seal populations. Before we can construct a trend, however, we first require a
baseline. Several studies have contributed estimates of the distribution and abundance of
ice-associated seal species in the Arctic using aerial surveys (e.g., [@bengtson2005pb],
[@conn2014mee], and [@verhoef2014sm]). Such studies are conducted over very large areas
and estimation of absolute abundance requires making inference about numerous nuisance
processes affecting the observation of seals on ice. This includes availability (only
seals basking on ice are available to be counted), detection probability (observers or
automated detection systems may miss some seals on ice), species misclassification, and
possible disturbance of seals by aircraft [@conn2014mee; @verhoef2014sm].

In this study, we use data from 15 years of bio-logger records in a comprehensive
investigation of haul-out behavior by 3 ice-associated seals in the Bering and Chukchi
Seas. We focused our effort on bearded, spotted, and ribbon seals and do not include
ringed seals in the analysis. Ringed seals are unique in their use of snow lairs in the
spring months, and they warrant a separate study that includes additional environmental
covariates, such as the onset of snow melt and snow depth. Our goals are threefold. First,
we wish to establish baseline estimates for the chronology of haul-out behavior in the
critical spring season for each species across different age and sex classes. Second, we
wish to refine estimates of haul-out availability corrections for aerial surveys in order
to improve estimates of seal abundance. Previously estimated availability correction
factors (e.g., [@bengtson2005pb], [@conn2014mee], and [@verhoef2014sm]) accounted for
variables such as the time-of-day and day-of-year, but did not investigate meteorological
variables that have been shown to influence haul-out behavior of walruses [@reder2003pb;
@udevitz2009cjz]. Third, we explore the annual variability in haul-out timing and possible
linkage to dramatic changes in the availability and extent of seasonal sea ice between
2005 and 2020. Our work extends the scope of previous haul-out analyses, includes the
influence of meteorological variability and possible ramifications of non-representative
age-sex samples, and investigates the potential impact of a changing icescape on the
behavior of these species.

@@

```{r seal-hours-calendar, fig.height = 4.75, include=TRUE, fig.id = "dataCal", fig.cap="Distribution of haul-out behavior data from 1 March to 15 July for each species. The higher density of data from May and June coincides with peak molting when seals (and their attached bio-loggers) are more likely hauled out. Additionally, a majority of deployments started in April and May. Data were collated across all years between 2005 and 2020."}
ggplot(data = seal_hours_tbl) + 
  geom_point(aes(x = wkofyr, y = dayofwk, color = n, size = n), stroke = 0, shape = 15) +
  scale_size(range = c(1,4.35)) +
  scale_color_gradient(low = "#6baed6", high = "#08306b") +
  scale_x_continuous(breaks = c(13,18,23,27),
                     labels = c("April", "May", "June", "July")) +
  #scale_y_discrete(labels = c("Sat","","","Wed","","","Sun"))+
  guides(color = "none",
        size = guide_legend(override.aes = list(color = "#2171b5"),
                            title.position = "bottom", title.hjust = 0.5)
        ) +
  facet_grid(species ~ .) + 
  theme(
        axis.text.y = element_blank(),
        panel.grid = element_blank(),
        legend.position = "bottom") +
  labs(x = "", y = "day of week", color = "seal-hours",
       size = "seal hours of bio-logger data") +
  ggtitle("Haul-out behavior data is concentrated in the months of May and June",
          subtitle = "data presented in seal-hours collated across years (2005-2020)")
```

@@

```{r examplePlot, include=TRUE, fig.id="examplePlot", fig.cap="Haul-out behavior observations recorded by a bio-logger deployment on a ribbon seal over two years during the months of March, April, May, and June. Empty areas of the plot represent missing data that was not successfully transmitted from the tag"}
d <- dat.sf %>% 
  filter(speno == "HF2016_1002") %>% 
  dplyr::select(speno, haulout_dt, percent_dry) %>% 
  arrange(haulout_dt) %>% st_drop_geometry() %>% 
  mutate(year = lubridate::year(haulout_dt),
         month = lubridate::month(haulout_dt,label=TRUE),
         day = lubridate::day(haulout_dt),
         hour = lubridate::hour(haulout_dt))

p <- ggplot(d,aes(day,hour,fill=percent_dry))+
  geom_tile(color= "white",size=0) + 
  scale_fill_distiller(palette = "YlGnBu",
                       name = "hourly percent dry", aesthetics = "fill",
    guide = guide_colorbar(title.position = "bottom", barwidth = 15,
                           barheight = 0.5, title.hjust = 0.5)
    )
  # scale_fill_viridis_c(name = "hourly percent dry", aesthetics = "fill",
  #   guide = guide_colorbar(title.position = "bottom", barwidth = 15,
  #                          barheight = 0.5, title.hjust = 0.5)
  # )
p <- p + facet_grid(year~month)
#p <- p + scale_y_continuous(trans = "reverse")
p <- p + scale_x_continuous(breaks = c(1,10,20,30))
p <- p + theme(legend.position = "bottom") +
  theme(strip.background = element_rect(colour="white")) +
  theme(axis.ticks=element_blank()) +
  ggtitle("Observation records are incomplete because not all data is transmitted",
          subtitle = "percent-dry data for a single ribbon seal across two years") +
  xlab("day of month") + ylab("hour (UTC)")
p
```

Of key interest in this study is the relationship between haul-out behavior and a variety
of environmental covariates that vary with space and time. We explored the use of a
movement model to predict locations at specific times. However, the sparse nature of data
from some tags, especially those mounted to the rear flipper, resulted in poor modeling
performance or convergence issues. For this study, we calculated a weighted average daily
location where the inverse of the estimated Argos location error was used for the weight.
Any days where haul-out observations were present without any location data were filled
from the last calculated weighted average daily location, and any days where the location
intersected with land were removed from the dataset. Figure \@ref(fig:dataMap) shows the
spatial distribution of available haul-out behavior data for each species across the study
area.

@@

Models for all species include the following effects: day-of-year, solar hour,
temperature, wind speed, barometric pressure, precipitation, wind chill (represented by a
wind $\times$ temperature interaction [@udevitz2009cjz]), and day-of-year and time of day
interactions to permit diurnal patterns to change throughout the year. Spotted and ribbon
seal models included age-sex class and interactions between day-of-year and age-sex class,
but we omitted these from bearded seal models due to poor representation of age-sex
classes (Table \@ref(tab:tblDeploy)). Bearded seal models included a latitudinal effect
(and an interaction with day-of-year), since bearded seals occupy a substantial range and
we were interested in possible differences in the timing of haul-out behavior along a
latitudinal gradient. We omitted the latitudinal effect from ribbon and spotted seal
models since, during the spring, these species are most prevalent near the southern ice
edge in the Bering Sea [@conn2014mee].

@@

```{r tblCovariates, include=TRUE, tab.id = "tblCovariates", tab.cap = "Explanatory covariates used in analyses of binary haul-out records for ice-associated seals."}
covariate_tbl <- tibble::tribble(
  ~Covariate, ~Type, ~Source, ~Description,
  "Age-sex class",   "Categorical", "Field Assessment", 
  "young-of-the-year, subadult, adult male and adult female",
  "Hour", "Continuous; Fourier basis", "Bio-logger", 
  "local solar hour using 6 variables of a Fourier-series basis",
  "Day",   "Continuous", "Bio-logger", 
  "linear, quadratic, and cubic effects of day-of-year",
  "Precip", "Continuous", "NARR", 
  "convective precipitation (kg/m2)",
  "Pressure", "Continuous", "NARR", 
  "atmospheric pressure at sea level (kPa)",
  "Temp", "Continuous", "NARR",
  "air temperatures at 2m above the earth’s surface",
  "Wind", "Continuous", "NARR", 
  "northerly and easterly vector components for wind converted into a single wind speed via the Euclidean norm",
  "Northing", "Continuous", "Bio-logger","latitude divided by the mean latitude across all locations (for bearded seals only)"
)

@@

Haul-out records for seals are often characterized by multiple hours spent basking on ice
alternating with long periods of foraging (Figure \@ref(fig:examplePlot)). Commonly used
statistical models for binary data (e.g. logistic regression) assume independence among
responses, an assumption that is clearly violated if hourly responses are modeled. Any
analysis that ignores temporal dependence in responses will thus have overstated
precision.

@@

We followed Ver Hoef et al. [-@verhoef2014sm] in using linear, quadratic, and cubic
effects of day-of-year to represent seasonal changes in behavior. However, unlike previous
models for harbor seals [@london2012] and ice-associated seals [@verhoef2014sm], which
treated hour-of-day as a 24-level categorical variable to capture diurnal cycles, we
adopted a continuous formulation based on Fourier series that provides a flexible model
while preserving the inherent circularity needed for time-of-day effects (i.e., hour 0
should be equal to hour 24). It also represents hour-of-day with 6 variables, which is a
considerable reduction when compared to a 24 variable model, especially when fitting
models with interactions between hour-of-day and other variables (e.g., age-sex class,
day-of-year). According to this approach, we used the following specification for
hour-of-day effects:

$$
H_t=\sum_{i=1}^{3}{\alpha_{i1}cos{\left(\frac{\pi t}{2\left(i+1\right)}\right)}}+\alpha_{i2}sin{\left(\frac{\pi t}{2\left(i+1\right)}\right)}1
$$

where $H_t$ gives the effect for solar hour $t$ and $α_{ij}$ for $i$ = 1, 2, or 3 and $j$
= 1 or 2 are estimated parameters (regression coefficients).

For the second set of models examining inter-annual variation in sea ice use, we fitted
models with year by day-of-year interactions. However, in this case we only included
year$\times$day and year$\times$day2, omitting the main effects of year as well as
year$\times$day3 interactions because models with the latter effects were numerically
unstable. However, the modeled interactions were sufficient to allow shifts in haul-out
distribution, as one can show mathematically that a simple horizontal shift in timing of
haul-out distributions does not affect the main effects or cubic terms in a polynomial
regression model.

@@

A typical model fitting exercise would also include a model selection process. However,
AIC (and similar criteria) is not suitable when using pseudo-likelihoods, because
pseudo-data generated in the model fitting process [@verhoef2010cs] differ between models
[@teneyck2017]. After fitting GLMPM models, we instead used "type III" $F$-tests to
calculate $p$-values [@verhoef2010cs] to evaluate model performance and important terms. We
also produced predictions of haul-out behavior as a function of influential predictors
(e.g. solar hour, day-of-year, age-sex). Weather covariates for these predictions were
based on daily or hourly smoothed weather covariate values across the study region. Such
predictions were then used to develop haul-out probability surfaces, explore marginal
effects of weather covariates, and determine annual peaks in haul-out activity. The timing
of peak haul-out behavior was further used to regress against the annual maximum sea ice
extent in the study region.

@@

```{r ribbonHOCal, fig.height = 5.5, include=TRUE, fig.id = "ribbonHOCal", fig.cap="Predicted haul-out probability of ribbon seals from 1 March to 15 July for each age and sex class used in the model."}
plot_df <- ribbon_newdata %>%
  mutate(date = lubridate::as_date(yday, origin = "2015-01-01"),
         month = lubridate::month(date,label=TRUE),
         day = lubridate::day(date)) %>%
  filter(month != "Aug")

p <- ggplot(plot_df,aes(day,solar_hour,fill=ho_prob))+
  geom_tile(color= "white",size = 0) +
  scale_fill_viridis_c(name = "haul-out probability", aesthetics = "fill",
                       guide = guide_colorbar(title.position = "bottom", barwidth = 15,
                                              barheight = 0.5, title.hjust = 0.5)
  )
p <- p + facet_grid(age_sex~month)
p <- p + scale_x_continuous(breaks = c(1,10,20,30))
p <- p + theme(legend.position = "bottom") +
  theme(strip.background = element_rect(colour="white")) +
  theme(axis.ticks=element_blank()) +
  xlab("day of month") + ylab("local solar hour") +
  ggtitle("Ribbon seals rest on sea ice in late spring and around solar noon.")
p
```

```{r}
ribbon_type3 <- tibble(
  fit_ribbon$typeIII.hypoth
) %>% 
  mutate(Prob.F = ifelse(Prob.F<0.001, "<0.01",Prob.F),
         Effect = janitor::make_clean_names(Effect)) %>% 
  column_to_rownames(var = "Effect") 
ribbon_type3 <- setNames(split(ribbon_type3, 
                               seq(nrow(ribbon_type3))), 
                         rownames(ribbon_type3))

spotted_type3 <- tibble(
  fit_spotted$typeIII.hypoth
) %>% 
  mutate(Prob.F = ifelse(Prob.F<0.001, "<0.01",Prob.F),
         Effect = janitor::make_clean_names(Effect)) %>%  
  column_to_rownames(var = "Effect")  
spotted_type3 <- setNames(split(spotted_type3, 
                               seq(nrow(spotted_type3))), 
                         rownames(spotted_type3))
bearded_type3 <- tibble(
  fit_bearded$typeIII.hypoth
) %>% 
  mutate(Prob.F = ifelse(Prob.F<0.001, "<0.01",Prob.F),
         Effect = janitor::make_clean_names(Effect)) %>%  
  column_to_rownames(var = "Effect")  
bearded_type3 <- setNames(split(bearded_type3, 
                               seq(nrow(bearded_type3))), 
                         rownames(bearded_type3))
```

>>>>

```{r setup, include = FALSE}
library(conflicted)
library(tidyverse)
conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")
library(lubridate)
conflict_prefer("month","lubridate")
library(glue)
library(scales)
library(DBI)
library(dbplyr)
library(sf)
library(flextable)
library(patchwork)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggspatial)
library(ggsci)
library(ggrepel)
library(gghighlight)
library(ggtext)
library(glmmLDTS)
library(mgcv)
library(ragg)

@@

The global climate crisis is causing considerable reduction in Arctic sea ice extent,
volume, and seasonal presence [@kwok2018erl, @meier2014rg, @wang2017dsrpitsio]. These
changes have ripple effects on Arctic ecosystems, taxa, and the human communities who live
in the region [@huntington2020ncc]. Such disruptions are a particular cause of concern for
ice-associated seals (ringed (*Phoca hispida*), bearded (*Erignathus barbatus*), spotted
(*Phoca largha*), and ribbon seals (*Histriophoca fasciata*)) that depend on spring and
early summer sea ice (March-June) in the Bering and Chukchi seas as a platform for
important functions, such as pupping, nursing, breeding, and molting [@boveng2009a;
@boveng2013; @cameron2010a; @kelly2010]. Limited data and large knowledge gaps prevent our
ability to make definitive predictions about the ultimate effects of changes in sea ice on
the behavior, abundance, and distribution of these seals. Knowledge about biological
constraints on the phenology of reproductive and molting behavior is generally lacking, so
it is difficult to predict the readiness with which ice-associated seal species will be
able to adapt to future changes (e.g., by adjusting pupping or molting schedules to
earlier dates or different locales). Additionally, trends in abundance of these species
are largely unknown, so it is difficult to say what effect, if any, declines in sea ice
habitat have had, or will have, on seal densities.

For ice-associated seals, haul out is tightly linked with sea ice. Ribbon seals rest out
of the water almost exclusively on sea ice and are mostly pelagic outside the spring
pupping, breeding, and molting season [@burns1981]. While spotted and bearded seals rest
on beaches during ice-free months, they strongly prefer sea ice as a resting platform
during the spring and early summer [@burns2009]. The remote nature of the Bering and
Chukchi seas mean direct observation of seal behavior is limited. Bio-logging devices are,
thus, especially useful tools for collecting key information on movement and haul-out
behavior for these species. Records of time spent out of the water provide valuable data
for identifying covariates that explain variation in haul-out behavior. For instance, in
the Antarctic, Bengtson and Cameron [-@bengtson2004pb] relied on bio-logging data to
demonstrate greater haul-out propensity in juvenile crabeater seals (*Lobodon
carcinophaga*) than adults, with highest probabilities in February and at times close to
solar noon. In the Arctic, Von Duyke et al. [-@vonduyke2020ee] demonstrated seasonal
changes between diurnal and nocturnal haul-out behavior of ringed seals. Bengtson et al.
[-@bengtson2005pb] documented a higher propensity for ringed seal basking near solar noon,
as did Ver Hoef et al. [-@verhoef2014sm] in an analysis of bearded , ribbon , and spotted
seals using much larger sample sizes. Olnes et al. [-@olnes2020pb] showed that the
proportion of time bearded seals spent hauled out progressively increased through spring
and summer. And, Ver Hoef et al. [-@verhoef2014sm] found haul-out probabilities increased
gradually starting in March and peaking in May and June for bearded, ribbon, and spotted
seals. However, their primary focus was to develop a correction factor for estimating seal
abundance, and like a subsequent analysis by Conn et al. [-@conn2014mee], they did not
attempt to separately analyze haul-out schedules for different age and sex classes
(although initial models with sex and age as fixed effects were insignificant at p = 0.05
[@verhoef2014sm]).

Several authors have used logistic regression-style analyses to estimate the probability
of sea ice use from data garnered from bio-loggers. In these models, haul-out
probabilities were expressed as a function of predictive covariates, such as time of day,
day-of-year, sex, age class, and environmental conditions (e.g., [@reder2003pb],
[@bengtson2004pb], [@bengtson2005pb], [@udevitz2009cjz], and [@verhoef2014sm]). However,
sample sizes have often been low and insufficient to permit strong inference about life
history and/or seasonal variation in haul-out probabilities. For instance, Bengtson and
Cameron's [-@bengtson2004pb] study included 5 adult and 2 juvenile crabeater seals, while
Bengtson et al.'s [-@bengtson2005pb] study was based on 6 telemetered ringed seals in the
Chukchi Sea. These studies were often further limited by logistical constraints on tagging
fieldwork and the attachment duration or operational life of satellite tags.

Ultimately, knowledge of trends in phenology and abundance (or life history surrogates
such as survival and recruitment) will be necessary to make credible quantitative
predictions about the effects of the climate crisis on the abundance and distribution of
Arctic seal populations. Before we can construct a trend, however, we first require a
baseline. Several studies have contributed estimates of the distribution and abundance of
ice-associated seal species in the Arctic using aerial surveys (e.g., [@bengtson2005pb],
[@conn2014mee], and [@verhoef2014sm]). Such studies were conducted over very large areas
and estimation of absolute abundance required making inference about numerous nuisance
processes affecting the observation of seals on ice. This included availability (only
seals basking on ice were available to be counted), detection probability (observers or
automated detection systems may have missed some seals on ice), species misclassification,
and possible disturbance of seals by aircraft [@conn2014mee; @verhoef2014sm].

In this study, we used data from 15 years of bio-logger records in a comprehensive
investigation of haul-out behavior by 3 species of ice-associated seals in the Bering and
Chukchi seas. We focused our effort on bearded, spotted, and ribbon seals and did not
include ringed seals in the analysis. Ringed seals are unique in their use of snow lairs
in the spring months, and they warrant a separate study that includes additional
environmental covariates, such as the onset of snow melt and snow depth. Our goals are
threefold. First, we wish to establish baseline estimates for the chronology of haul-out
behavior in the critical spring season for each species across different age and sex
classes. Second, we wish to refine estimates of haul-out availability corrections for
aerial surveys in order to improve estimates of seal abundance. Previously estimated
availability correction factors (e.g., [@bengtson2005pb], [@conn2014mee], and
[@verhoef2014sm]) accounted for variables such as the time-of-day and day-of-year, but did
not investigate meteorological variables that have been shown to influence haul-out
behavior of walruses [@reder2003pb; @udevitz2009cjz]. Third, we explored the annual
variability in haul-out timing and possible linkage to dramatic changes in the
availability and extent of seasonal sea ice between 2005 and 2020. Our work extends the
scope of previous haul-out analyses, includes the influence of meteorological variability,
and investigates the potential impact of a changing icescape on the behavior of these
species.

@@

```{r seal-hours-calendar, include=TRUE, fig.height = 4.75, fig.id = "dataCal", fig.cap="Distribution of haul-out behavior data from 1 March to 15 July for each species. The higher density of data from May and June coincides with peak molting when seals (and their attached bio-loggers) are more likely hauled out. Additionally, a majority of deployments started in April and May. Data were collated across all years between 2005 and 2020."}
ggplot(data = seal_hours_tbl) + 
  geom_point(aes(x = wkofyr, y = dayofwk, color = n, size = n), stroke = 0, shape = 15) +
  scale_size(range = c(1,4.35)) +
  scale_color_gradient(low = "#6baed6", high = "#08306b") +
  scale_x_continuous(breaks = c(13,18,23,27),
                     labels = c("April", "May", "June", "July")) +
  #scale_y_discrete(labels = c("Sat","","","Wed","","","Sun"))+
  guides(color = "none",
        size = guide_legend(override.aes = list(color = "#2171b5"),
                            title.position = "bottom", title.hjust = 0.5)
        ) +
  facet_grid(species ~ .) + 
  theme(
        axis.text.y = element_blank(),
        panel.grid = element_blank(),
        legend.position = "bottom") +
  labs(x = "", y = "day of week", color = "seal-hours",
       size = "seal hours of bio-logger data") +
  ggtitle("Haul-out behavior data is concentrated in the months of May and June",
          subtitle = "data presented in seal-hours collated across years (2005-2020)")
```

@@

```{r examplePlot, include=TRUE, fig.id="examplePlot", fig.cap="Haul-out behavior observations recorded by a bio-logger deployment on a ribbon seal over two years during the months of March, April, May, and June. Empty areas of the plot represent missing data that were not successfully transmitted from the tag"}
d <- dat.sf %>% 
  filter(speno == "HF2016_1002") %>% 
  dplyr::select(speno, haulout_dt, percent_dry) %>% 
  arrange(haulout_dt) %>% st_drop_geometry() %>% 
  mutate(year = lubridate::year(haulout_dt),
         month = lubridate::month(haulout_dt,label=TRUE),
         day = lubridate::day(haulout_dt),
         hour = lubridate::hour(haulout_dt))

p <- ggplot(d,aes(day,hour,fill=percent_dry))+
  geom_tile(color= "white",size=0) + 
  scale_fill_distiller(palette = "YlGnBu",
                       name = "hourly percent dry", aesthetics = "fill",
    guide = guide_colorbar(title.position = "bottom", barwidth = 15,
                           barheight = 0.5, title.hjust = 0.5)
    )
  # scale_fill_viridis_c(name = "hourly percent dry", aesthetics = "fill",
  #   guide = guide_colorbar(title.position = "bottom", barwidth = 15,
  #                          barheight = 0.5, title.hjust = 0.5)
  # )
p <- p + facet_grid(year~month)
#p <- p + scale_y_continuous(trans = "reverse")
p <- p + scale_x_continuous(breaks = c(1,10,20,30))
p <- p + theme(legend.position = "bottom") +
  theme(strip.background = element_rect(colour="white")) +
  theme(axis.ticks=element_blank()) +
  ggtitle("Observation records are incomplete because not all data are transmitted",
          subtitle = "percent-dry data for a single ribbon seal across two years") +
  xlab("day of month") + ylab("hour (UTC)")
p
```

Of key interest in this study is the relationship between haul-out behavior and a variety
of environmental covariates that vary with space and time. We explored the use of a
movement model to predict locations at specific times. However, the sparse nature of data
from some tags, especially those mounted to the rear flipper, resulted in poor modeling
performance or convergence issues. For this study, we calculated a weighted average daily
location where the inverse of the estimated Argos location error was used for the weight.
Any days where haul-out observations were present without any location data were filled
from the last calculated weighted average daily location, and any days where the location
intersected with land were removed from the dataset. Figure \@ref(fig:dataMap) shows the
spatial distribution of determined locations and available haul-out behavior data for each
species across the study area.

@@

Models for all species include the following effects: day-of-year, solar hour,
temperature, wind speed, barometric pressure, precipitation, wind chill (represented by a
_wind:temperature_ interaction [@udevitz2009cjz]), and day-of-year and time of day
interactions to permit diurnal patterns to change throughout the year. Spotted and ribbon
seal models included age-sex class and interactions between day-of-year and age-sex class,
but we omitted these from bearded seal models due to poor representation of age-sex
classes (Table \@ref(tab:tblDeploy)). Bearded seal models included a latitudinal effect
(and an interaction with day-of-year), since bearded seals occupy a substantial range and
we were interested in possible differences in the timing of haul-out behavior along a
latitudinal gradient. We omitted the latitudinal effect from ribbon and spotted seal
models since, during the spring, these species are most prevalent near the southern ice
edge in the Bering Sea [@conn2014mee].

@@

```{r tblCovariates, include=TRUE, tab.id = "tblCovariates", tab.cap = "Explanatory covariates used in analyses of binary haul-out records for ice-associated seals.  Note that we also considered select interactions (see article text) between these primary covariates.  For instance, wind chill was represented by the interaction _temperature:wind_."}
covariate_tbl <- tibble::tribble(
  ~Covariate, ~Type, ~Source, ~Description,
  "Age-sex class",   "Categorical", "Field Assessment", 
  "young-of-the-year, subadult, adult male and adult female",
  "Hour", "Continuous; Fourier basis", "Bio-logger", 
  "local solar hour using 6 variables of a Fourier-series basis",
  "Day",   "Continuous", "Bio-logger", 
  "linear, quadratic, and cubic effects of day-of-year",
  "Precip", "Continuous", "NARR", 
  "convective precipitation (kg/m2)",
  "Pressure", "Continuous", "NARR", 
  "atmospheric pressure at sea level (kPa)",
  "Temp", "Continuous", "NARR",
  "air temperatures at 2m above the earth’s surface",
  "Wind", "Continuous", "NARR", 
  "northerly and easterly vector components for wind converted into a single wind speed via the Euclidean norm",
  "Northing", "Continuous", "Bio-logger","latitude divided by the mean latitude across all locations (for bearded seals only)"
)

@@

Haul-out records for seals are often characterized by multiple hours spent basking on ice
alternating with long periods of foraging (Figure \@ref(fig:examplePlot)). Commonly used
statistical models for binary data (e.g. logistic regression) assume independence among
responses, an assumption that is clearly violated if hourly responses are modeled. Any
analysis that ignores temporal dependence in responses will thus have overstated precision
[@betts2006].

@@

We followed Ver Hoef et al. [-@verhoef2014sm] in using linear, quadratic, and cubic
effects of day-of-year to represent seasonal changes in behavior. However, unlike previous
models for harbor seals [@london2012] and ice-associated seals [@verhoef2014sm], which
treated hour-of-day as a 24-level categorical variable to capture diurnal cycles, we
adopted a continuous formulation based on Fourier series that provides a flexible model
while preserving the inherent circularity needed for time-of-day effects (i.e., hour 0
should be equal to hour 24). It also represents hour-of-day with 6 parameters, which is a
considerable reduction when compared to a 24 parameter variable, especially when fitting
models with interactions between hour-of-day and other variables (e.g., age-sex class,
day-of-year). According to this approach, we used the following specification for
hour-of-day effects:

$$
H_t=\alpha_1 \cos(\frac{\pi t}{4}) + \alpha_2 \sin(\frac{\pi t}{4}) + \alpha_3 \cos(\frac{\pi t}{6}) + \alpha_4 \sin(\frac{\pi t}{6}) + \alpha_5 \cos(\frac{\pi t}{12}) + \alpha_6 \sin(\frac{\pi t}{12})
$$ 
where $H_t$ gives the effect for solar hour $t$ and $α_{i}$ are estimated parameters
(regression coefficients).

For the second set of models examining inter-annual variation in sea ice use, we fitted
models with year by day-of-year interactions. However, in this case we only included
_year:day_ and _year:day2_, omitting the main effects of year as well as
_year:day3_ interactions because models with the latter effects were numerically
unstable. However, the modeled interactions were sufficient to allow shifts in haul-out
distribution, as one can show mathematically that a simple horizontal shift in timing of
haul-out distributions does not affect the main effects or cubic terms in a polynomial
regression model.

@@

A typical model fitting exercise would also include a model selection process. However,
AIC (and similar criteria) is not suitable when using pseudo-likelihoods, because
pseudo-data generated in the model fitting process [@verhoef2010cs] differ between models
[@teneyck2017]. After fitting GLMPM models, we instead used "type III" $F$-tests to
calculate $p$-values [@verhoef2010cs] to evaluate model performance and important terms.
We also produced predictions of haul-out behavior as a function of influential predictors
(e.g. solar hour, day-of-year, age-sex). Weather covariates for these predictions were
based on daily or hourly smoothed weather covariate values across the study region. Such
predictions were then used to develop haul-out probability surfaces, explore conditional
effects of weather covariates, and determine annual peaks in haul-out activity. The timing
of peak haul-out behavior was further used to regress against the annual maximum sea ice
extent in the study region.

@@

```{r ribbonHOCal, fig.height = 5.5, include=TRUE, fig.id = "ribbonHOCal", fig.cap="Predicted haul-out probability of ribbon seals from 1 March to 15 July for each age and sex class used in the model."}
plot_df <- ribbon_newdata %>%
  mutate(date = lubridate::as_date(yday, origin = "2015-01-01"),
         month = lubridate::month(date,label = TRUE),
         day = lubridate::day(date)) %>%
  filter(month != "Aug")

p <- ggplot(plot_df, aes(day,solar_hour,fill = ho_prob)) +
  geom_tile(color = "white",size = 0) +
  scale_fill_viridis_c(name = "haul-out probability", aesthetics = "fill",
                       guide = guide_colorbar(title.position = "bottom", barwidth = 15,
                                              barheight = 0.5, title.hjust = 0.5)
  )
p <- p + facet_grid(age_sex~month)
p <- p + scale_x_continuous(breaks = c(1,10,20,30))
p <- p + theme(legend.position = "bottom") +
  theme(strip.background = element_rect(colour = "white")) +
  theme(axis.ticks = element_blank()) +
  xlab("day of month") + ylab("local solar hour") +
  ggtitle("Ribbon seals rest on sea ice in late spring and around solar noon.")
p
```

```{r}
ribbon_type3 <- tibble(
  fit_ribbon$typeIII.hypoth
) %>% 
  mutate(Prob.F = scales::pvalue(Prob.F),
         Effect = janitor::make_clean_names(Effect)) %>% 
  column_to_rownames(var = "Effect") 
ribbon_type3 <- setNames(split(ribbon_type3, 
                               seq(nrow(ribbon_type3))), 
                         rownames(ribbon_type3))

spotted_type3 <- tibble(
  fit_spotted$typeIII.hypoth
) %>% 
  mutate(Prob.F = scales::pvalue(Prob.F),
         Effect = janitor::make_clean_names(Effect)) %>%  
  column_to_rownames(var = "Effect")  
spotted_type3 <- setNames(split(spotted_type3, 
                               seq(nrow(spotted_type3))), 
                         rownames(spotted_type3))
bearded_type3 <- tibble(
  fit_bearded$typeIII.hypoth
) %>% 
  mutate(Prob.F = scales::pvalue(Prob.F),
         Effect = janitor::make_clean_names(Effect)) %>%  
  column_to_rownames(var = "Effect")  
bearded_type3 <- setNames(split(bearded_type3, 
                               seq(nrow(bearded_type3))), 
                         rownames(bearded_type3))
```
